NURM-tool: a quantitative reproducibility measurement tool to identify numerical differences in pipelines

### 2019-05-02 paper.tex
add clustering section

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} more specifically on some of 
the main neuroimaging pipelines including FSL, CIVET and Freesurfer 
pipelines. The results of comparisons indicate a significant differences 
between analysis results of the same experiment on different operating 
system. \note{It should be noted that the binary 
differences are measured between results of two analyses in this paper, which is 
calculated by the checksum method. }
The operating system is not the only part of the computational 
environment that may hamper reproducibility. Many other computational 
factors including hardware configuration, software version, and 
parallelization techniques are reported as the main influential 
computing infrastructure on reproducibility.

@@

We present a Numerical Reproducibility Measurement tool (NURM-tool) to 
measure numerical reproducibility of the pipelines across operating 
systems automatically (see Figure~\ref{fig:overview}). We build the 
containerized version of pipelines for each operating system using 
docker technology which encapsulated whole the pipeline dependencies. 
The docker images consists of an specific computing environment such as 
operating system version, hardware configuration, software version, 
etc. Furthermore, to simplify computational reproducibility of the 
analysis, we execute docker images using Boutiques framework based on 
the command-line descriptions~\cite{glatard2017boutiques}. The 
Boutiques descriptor specifies the parameters and implementation of the 
experiment, and the invocation schema describes the parameter values 
such as input/output data path.

NURM-tool produces the error JSON schema in which recognizes output files 
with errors using the checksum comparison of the results. 
We consider that a file contains errors if 
its binary comparison is different, otherwise file has no error. Also, 
we use \reprozip tool to capture the provenance information. 
Subsequently, using a iterative approach, we classify processes into 
different categories based on the state of read/write files 
with/without error. NURM-tool identifies relevant processes in the 
pipeline creating, propagating, or removing errors. We will explain 
each steps in detail in the following sections.

@@

Depending on the variety of computing environments, different  
process tree can be produced. In particular, different analysis 
software package can construct different process trees for the 
similar experiments because of following different methods. We used a 
fixed analysis software package in different conditions, however, we 
found that different subjects may produce a different process trees.

The challenging problem is that of comparing various process trees with 
different topological structures from different subjects. To address 
this issue, we categorize subjects into different types based on the 
distances, topologically differences, between corresponding process 
trees. 
We cluster subjects into various types using the bottom-up 
hierarchical clustering approach. This approach groups together the 
subjects with similar distance, less than a threshold.

@@

The clustering algorithm starts by finding the two subjects that are 
closest to each other on the basis of edit distance method and form the 
first cluster between these two subjects. The next step join this 
cluster to the next nearest cluster or subject which in turn results in 
another cluster. This process continues until all the subjects are 
joined together to form one big cluster or terminates when the distances 
are less than the threshold. The threshold is defined as the minimum 
distance required to separate clusters.

\subsection{Graph analysis}

NURM-tool classifies the processes into four 
categories based on the process graph and the error JSON file 
mentioned in the previous sections:
\begin{enumerate}
\item Processes that read files that do not have errors and write files 
that do not have errors are \emph{transparent} (Figure~\ref{fig:processes}.a).
\item Processes that read files 
that do not have any error but write files that have errors 
\emph{create} errors in the pipeline (Figure~\ref{fig:processes}.b).
\item Processes that read files 
that have errors and write files that do not have errors \emph{remove} 
errors from the pipeline (Figure~\ref{fig:processes}.c).
\item Processes that read files that have errors and write files that also have errors are 
\emph{unknown} (Figure~\ref{fig:processes}.d).
\end{enumerate}

\begin{figure}%\centering
\centering
    \begin{subfigure}{0.4\linewidth}
        \includegraphics[scale=0.34]{images/green.png}
        \caption{Transparent}
        \label{fig:green}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\linewidth}
    \includegraphics[scale=0.34]{images/red.png}
    \caption{Creates error}
    \label{fig:red}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\linewidth}
    \includegraphics[scale=0.34]{images/blue.png}
    \caption{Removes error}
    \label{fig:blue}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\linewidth}
    \includegraphics[scale=0.34]{images/yellow.png}
    \caption{Unknown}
    \label{fig:yellow}
\end{subfigure}
    \caption{Different type of classified process based on the input/output files.
  Dashed edges refer to the file dependencies between processes A and B 
  if a file written by A is read by B. Solid black edges refer to the 
  relationship between parent and child processes.}
    \label{fig:processes}
\end{figure}

@@

Furthermore, cycles may be present in the process graph in case a file 
was written by more than one process. We remove such cycles by removing 
file edges between processes A and B when A's process creation 
timestamp is posterior to B's or when A=B. Indeed, such edges cannot 
happen in practice unless A and B were running concurrently, which we 
assume is not the case (we also checked that on the workflow we 
tested).

\subsubsection{Unknown process} 
Another challenge is to classify processes that 
read files that have errors and write files that also have errors. In 
such situations, it is not possible to determine from the pipeline 
results whether the process created errors, removed some errors, or was 
transparent. We label such processes \emph{unknown}.
To address this issue, we developed an iterative approach that 
consists of the following steps:

\begin{enumerate}
  \item Run the pipeline in conditions 1 and 2; classify the
    processes as \emph{transparent}, \emph{create errors},
    \emph{remove errors} or \emph{unknown}.
  \item If there are \emph{unknown} processes then: replace each 
  process P that create errors by a process Q that copies the results 
  produced by process P in condition 1 to the pipeline output in 
  condition 2. 
  \item Repeat steps 1 and 2 until there is no \emph{unknown} process left.
\end{enumerate}

The replacement of process P at step 2 is done by replacing P with a 
custom script produced by step 1. This custom script copies the results 
obtained in condition 1 if it is invoked with the arguments of a 
process that created errors, and calls P's original executable 
otherwise. The replacement is done through the PATH variable, as 
before. This algorithm converges to a process graph without any 
\emph{unknown} process after a finite number of iterations.

@@

Figure~\ref{fig:iterations} illustrates our iterative classification 
process for the example in Figure~\ref{fig:simple_script}. At every 
step, processes that created errors are shown in red, processes that 
removed errors are in blue, processes that propagated errors are in 
yellow, and other processes (transparent processes) are in green. As in 
Figure~\ref{fig:simple_script}, plain black edges represent the process 
tree and dashed edges represent file dependencies: green edges 
represent files with no errors, while red edges represent files with 
errors. Temporary files are not represented because they have been 
backed up as previously described.

The three steps in Figure~\ref{fig:iterations} correspond to the 
iterations of the classification algorithm. At step one, \texttt{2\#bet2} 
is classified as error creator (red) as it produced files with errors 
from files without error. \texttt{3\#fslstats} is classified as error 
remover (blue) as it produced files without errors from files with 
errors. \texttt{4\#fslmaths} and \texttt{5\#fslstats} are classified as 
unknown (yellow) as they produce files with errors from files with 
errors.

At step 2, the files produced by \texttt{2\#bet2} in the tested 
condition are replaced with the files produced by \texttt{2\#bet2} in 
the other condition. \texttt{3\#fslstats} is now classified as 
transparent, \texttt{4\#fslmaths} is now classified as error creator 
and \texttt{5\#fslstats} is still unknown.

@@

As a result of those 3 steps, the final process classification is: 
\texttt{2\#bet2} and \texttt{4\#fslmath} are error creators (red), 
\texttt{3\#fslstats} is error remover (blue) and the other processes 
are transparent.

@@

We randomly selected 10 subjects from the HCP data release S500 which 
are described in table~\ref{table:data}.

@@

\begin{table}
\centering
\begin{threeparttable}
\caption{An overview of the HCP data (Humman Connectome DB).}

\begin{tabular}{@{}llllll@{}}
\toprule
Subject & Release & Acquisition & Gender & Age      & Subj\_Type   \\ \midrule
103515  & Q1      & Q02         & F      & 26-30    & type1         \\
105216  & Q3      & Q03         & M      & 26-30    & type4         \\
103414  & Q2      & Q02         & F      & 22-25    & type1         \\
125525  & Q1      & Q01         & F      & 31-35    & type1         \\
142828  & Q1      & Q01         & M      & 31-35    & type1         \\
129533  & Q3      & Q04         & F      & 31-35    & type1         \\
103818  & Q1      & Q01         & F      & 31-35    & type3         \\
133928  & Q2      & Q03         & M      & 26-30    & type3         \\
148032  & Q3      & Q03         & F      & 31-35    & type2         \\
139637  & Q2      & Q03         & F      & 31-35    & type1         \\ \bottomrule
\end{tabular}
\begin{tablenotes}
      \small
      \item *Subj\_Type refers to the graph topology type of the subjects.
\end{tablenotes}
\end{threeparttable}
\label{table:data}
\end{table}

\note{Among our dataset, we identified 4 types of subjects with 
different numbers of T1w and T2w images. In addition, we verified that 
the process graphs were identical for all subjects of the same type in 
different versions of CentOS operating system. The remainder of the 
analysis is done separately for each subject type.}

@@

CBRAIN team. Compute Canada(Calcul Quebec).

@@

>>>>

In particular, the effect of operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} more specifically on some of 
the main neuroimaging pipelines including FSL, CIVET and Freesurfer 
pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating system, which indicate 
a significant disparity. 
The operating system is not the only part of the computational 
environment that may hamper reproducibility. Many other computational 
factors including hardware configuration, software version, and 
parallelization techniques are reported as the main influential 
computing infrastructure on reproducibility.

@@

We present NURM-tool to automatically
measure numerical reproducibility of the pipelines across operating 
systems (see Figure~\ref{fig:overview}). We build the 
containerized version of the pipelines for each operating system using 
docker technology which encapsulated whole the pipeline dependencies. 
The docker images consists of an specific computing environment such as 
operating system version, hardware configuration, software version, 
etc. Furthermore, to simplify computational reproducibility of the 
analysis, we execute docker images using Boutiques framework based on 
the command-line descriptions~\cite{glatard2017boutiques}. The 
Boutiques descriptor specifies the parameters and implementation of the 
experiment, and the invocation schema describes the parameter values 
such as input/output data path.

As shown in the left side of Figure~\ref{fig:overview}, NURM-tool clusters 
the input subjects into various types based on the similarity of their 
process trees which are achieved by \reprozip tool. The right side of this figure 
shows the iterative classification of the process based on the binary 
comparison of the outputs from various conditions.
NURM-tool produces the \textit{difference} JSON schema in which 
recognizes output files with differences using the checksum comparison 
of the results. We consider that a file contains difference if only 
its binary comparison shows difference. Also, 
we use \reprozip tool to capture the provenance information. 
Subsequently, using a iterative approach, we classify processes into 
different categories regarding the state of read/write files 
with/without differences. NURM-tool identifies relevant processes in the 
pipeline creating, propagating, or removing differences. 
All the steps are explained in detail in the following sections.

@@

Depending on the variety of computing environments, different  
process tree can be produced. In particular, different analysis 
software package can construct different process tree for the 
similar experiments because of following different methods. We used a 
fixed analysis software package in different conditions, however, we 
found that different subjects may produce a different process tree.

The challenging problem is that of comparing various process trees with 
different topological structures from different subjects. To address 
this issue, we categorize subjects into different types based on the 
distances, topologically differences, between corresponding process 
trees.

@@

We cluster subjects into various types using the bottom-up hierarchical 
clustering approach implemented in \textit{scipy} python package (see 
Algorithm~\ref{also:hclustering}). This approach groups together the 
subjects with similar distance, less than a threshold.

\begin{algorithm}[h!]
\caption{Hierarchical clustering algorithm}
\label{algo:hclustering}
\begin{algorithmic}

  \State /* Input: a list of n trees; \texttt{T = [t${_1}$, t${_2}$,...,t${_n}$]}
  \State /* Output: a list of clusters; \texttt{C = [c${_1}$,c${_2}$,...,c${_k}$]}
  \For{\texttt{i=1} in \texttt{n}}
  \State \texttt{C${_i}$} = \texttt{T${_i}$} 
  \State \texttt{\# Create a cluster for each tree in input list; \texttt{agglomerative approach}}
  \EndFor
  \While{\texttt{len(C)} > \texttt{1}}  
  \For{\texttt{i=1} to \texttt{n}}
  \For{\texttt{j=i+1} to \texttt{n}}
  \State \texttt{D${_{ij}}$} = argmin\{\texttt{edit\_distance(C${_i}$, C${_j}$)}\}
  \State \texttt{\# Select two nearest clusters based on \texttt{nearest\_neighbors} method}
  \EndFor
  \EndFor
  \If{\texttt{D${_{ij}}$} > \texttt{thre}}
  \State Return \texttt{C}
  \Else {\State Merge \texttt{C${_i}$ and C${_j}$}}
  \EndIf
  \EndWhile
%  \State Function edit\_distance (C${_1}$, C${_2}$)
%  \State EndFunction
\end{algorithmic}
\end{algorithm}

The clustering algorithm starts by finding the two subjects that are 
closest to each other on the basis of edit distance method and form the 
new cluster between these two subjects. The next step join this 
cluster to the next nearest cluster or subject which in turn results in 
another cluster. This process continues until all the subjects are 
joined together to form one big cluster or terminates when the distances 
are less than the threshold. The threshold is defined as the minimum 
distance required to separate clusters.

\subsection{Tree analysis}

NURM-tool classifies the processes into four 
categories based on the process tree and the \textit{difference} JSON file 
mentioned in the previous sections:
\begin{enumerate}
\item Processes that read and write files that do not have differences 
are \emph{transparent} (Figure~\ref{fig:processes}.a).
\item Processes that read files 
that do not have any differences but write files that have differences 
\emph{create} differences in the pipeline (Figure~\ref{fig:processes}.b).
\item Processes that read files 
that have differences and write files that do not have differences \emph{remove} 
differences from the pipeline (Figure~\ref{fig:processes}.c).
\item Processes that read files that have differences and write files that 
also have differences are 
\emph{unknown} (Figure~\ref{fig:processes}.d).
\end{enumerate}

\begin{figure}%\centering
\centering
    \begin{subfigure}{0.4\linewidth}
        \includegraphics[scale=0.34]{images/green.png}
        \caption{Transparent}
        \label{fig:green}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\linewidth}
    \includegraphics[scale=0.34]{images/red.png}
    \caption{Creates differences}
    \label{fig:red}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\linewidth}
    \includegraphics[scale=0.34]{images/blue.png}
    \caption{Removes differences}
    \label{fig:blue}
    \end{subfigure}
    \hfill
    \begin{subfigure}{0.4\linewidth}
    \includegraphics[scale=0.34]{images/yellow.png}
    \caption{Unknown}
    \label{fig:yellow}
\end{subfigure}
    \caption{Different type of classified process based on the input/output files.
  Dashed edges refer to the file dependencies between processes A and B 
  if a file written by A is read by B. Solid black edges refer to the 
  relationship between parent and child processes.}
    \label{fig:processes}
\end{figure}

@@

Furthermore, cycles may be present in the process tree in case a file 
was written by more than one process. We remove such cycles by removing 
file edges between processes A and B when A's process creation 
timestamp is posterior to B's or when A=B. Indeed, such edges cannot 
happen in practice unless A and B were running concurrently, which we 
assume is not the case (we also checked that on the workflow we 
tested).

\subsubsection{Unknown process} 
Another challenge is to classify processes that 
read files that have differences and write files that also have differences. In 
such situations, it is not possible to determine from the pipeline 
results whether the process created differences, removed some differences, or was 
transparent. We label such processes \emph{unknown}.
To address this issue, we developed an iterative approach that 
consists of the following steps:

\begin{enumerate}
  \item Run the pipeline in conditions 1 and 2; classify the
    processes as \emph{transparent}, \emph{create differences},
    \emph{remove differences} or \emph{unknown}.
  \item If there are \emph{unknown} processes then: replace each 
  process P that create differences by a process Q that copies the results 
  produced by process P in condition 1 to the pipeline output in 
  condition 2. 
  \item Repeat steps 1 and 2 until there is no \emph{unknown} process left.
\end{enumerate}

The replacement of process P at step 2 is done by replacing P with a 
custom script produced by step 1. This custom script copies the results 
obtained in condition 1 if it is invoked with the arguments of a 
process that created differences, and calls P's original executable 
otherwise. The replacement is done through the PATH variable, as 
before. This algorithm converges to a process tree without any 
\emph{unknown} process after a finite number of iterations.

@@

Figure~\ref{fig:iterations} illustrates our iterative classification 
process for the example in Figure~\ref{fig:simple_script}. At every 
step, processes that created differences are shown in red, processes that 
removed differences are in blue, processes that propagated differences are in 
yellow, and other processes (transparent processes) are in green. As in 
Figure~\ref{fig:simple_script}, plain black edges represent the process 
tree and dashed edges represent file dependencies: green edges 
represent files with no differences, while red edges represent files with 
differences. Temporary files are not represented because they have been 
backed up as previously described.

The three steps in Figure~\ref{fig:iterations} correspond to the 
iterations of the classification algorithm. At step one, \texttt{2\#bet2} 
is classified as difference creator (red) as it produced files with differences 
from files without differences. \texttt{3\#fslstats} is classified as difference  
remover (blue) as it produced files without differences from files with 
differences. \texttt{4\#fslmaths} and \texttt{5\#fslstats} are classified as 
unknown (yellow) as they produce files with differences from files with 
differences.

At step 2, the files produced by \texttt{2\#bet2} in the tested 
condition are replaced with the files produced by \texttt{2\#bet2} in 
the other condition. \texttt{3\#fslstats} is now classified as 
transparent, \texttt{4\#fslmaths} is now classified as difference creator 
and \texttt{5\#fslstats} is still unknown.

@@

As a result of those 3 steps, the final process classification is: 
\texttt{2\#bet2} and \texttt{4\#fslmath} are difference creators (red), 
\texttt{3\#fslstats} is difference remover (blue) and the other processes 
are transparent.

@@

We randomly selected 100 subjects from the HCP data release S500 and then 
using the NURM-tool we cluster them into different groups. 
Figure~\ref{fig:subj-clusters} shows the result of data clustering. 
Among our dataset, we identified 4 types of subjects with 
different topology of process trees. In addition, we verified that 
the process trees were identical for all subjects of the same type in 
different versions of CentOS operating system. The remainder of the 
analysis is done separately for each subject type.

@@

%\begin{table}
%\centering
%\begin{threeparttable}
%\caption{An overview of the HCP data (Humman Connectome DB).}
%
%\begin{tabular}{@{}llllll@{}}
%\toprule
%Subject & Release & Acquisition & Gender & Age      & Subj\_Type   \\ \midrule
%103515  & Q1      & Q02         & F      & 26-30    & type1         \\
%105216  & Q3      & Q03         & M      & 26-30    & type4         \\
%103414  & Q2      & Q02         & F      & 22-25    & type1         \\
%125525  & Q1      & Q01         & F      & 31-35    & type1         \\
%142828  & Q1      & Q01         & M      & 31-35    & type1         \\
%129533  & Q3      & Q04         & F      & 31-35    & type1         \\
%103818  & Q1      & Q01         & F      & 31-35    & type3         \\
%133928  & Q2      & Q03         & M      & 26-30    & type3         \\
%148032  & Q3      & Q03         & F      & 31-35    & type2         \\
%139637  & Q2      & Q03         & F      & 31-35    & type1         \\ \bottomrule
%\end{tabular}
%\begin{tablenotes}
%      \small
%      \item *Subj\_Type refers to the tree topology type of the subjects.
%\end{tablenotes}
%\end{threeparttable}
%\label{table:data}
%\end{table}

@@

Data were provided by the Human Connectome Project, WU-Minn 
Consortium (Principal Investigators: David Van Essen and Kamil Ugurbil; 
1U54MH091657) funded by the 16 NIH Institutes and Centers that support 
the NIH Blueprint for Neuroscience Research; and by the McDonnell 
Center for Systems Neuroscience at Washington University.

\note{CBRAIN team. Compute Canada(Calcul Quebec).}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{images/hclusters}
  \caption{Different data types clustered among 100 subjects.}
  \label{fig:subj-clusters}
\end{figure}
